An event-sourced and distributed database with IPLD and materialized views

### 2019-09-23 draft.tex
update: ready for sharing?
Signed-off-by: Carson Farmer <carson.farmer@gmail.com>

One of the most exciting aspects of user-siloed data is the the ability to build data-driven apps and services while users remain in control of their own data. Other projects have identified app silo-ed data as a problem [ref], and some have identified user-siloed data as a solution \cite{sambraSolidPlatformDecentralized2016}. However, none so far have provided a sufficiently interoperable protocol for scalable storage and transmission of user-siloed application data. This is the key to building great applications on user-siloed data.

@@

\begin{definition} (Command query responsibility segregation). Command query responsibility segregation or CQRS is a design pattern whereby reads and writes are separated into different models, using commands to write data, and queries to read data [ref]. \end{definition}

@@

At a minimum, a node must link to its most immediate ancestor. However, links to older ancestors are often included as well to improve concurrency during traversal and verification [ref].

@@

The use of well-established design patterns from the CQRS+ES (and related) literature means the Threads protocol can provide the foundation for many higher-level concepts and technologies. For instance, the public application programming interface (API) for a Thread provides several features that you would expect when operating on a dataset or table within a database. Indeed, each Thread is defined by a unique ID (its Identity), and provides facilities for access control and permissions, networking, sharing, invites, and more.

Like many CQRS-based systems, the ``public'' API for Threads revolves around the concept of Views, which are used to optimize queries on the local event store. A key feature of Views is they are completely disposable, because they can be entirely rebuilt from the source event store\footnote{Though see section \ref{sec:SnapshotsCompaction} for a discussion of Snapshots, which may be used to speed up Views.}. In Threads, Views are updated by subscribing to specific Event(s), and applying a Reducer function that is used to update the View state (which may be \emph{materialized} to local persistent storage). This simple subscription system provides a flexible framework for building complex View-driven application logic.

\subsubsection{Default Views}\label{sec:DefaultViews}

The core Textile Threads interfaces are also exposed via interfaces compatible with \emph{existing} datastores and/or communications systems.
These simple, public-facing APIs will be comfortable to application developers looking to leverage highly-distributed systems that connect user-controlled data, with minimal configuration and maximum interoperability.
In the remainder of this section, we highlight a number of planned database abstractions to provide further utility to developers interested in building on Textile's Threads protocol.

\paragraph{Databases}

A key/value store built on top of Threads can map database \texttt{Put} operations into internal \texttt{KeyValueUpdated} events. These Events would then operate on a map-like view model, via an internal ``projection'' (reducer function). The key/value view could be persisted or kept in-memory, and would be directly queried via a corresponding \texttt{Get} operation. Because Textile uses ES patterns, a \texttt{Del} operation would simply lead to an internal \texttt{KeyValueDeleted} event, marking the key/value pair for removal from the key/value materialized view.

\begin{example}
\begin{minipage}{.45\textwidth}
\begin{lstlisting}
type TextileKVStore interface {
  Put(key string, value interface{}) error
  Get(key string) (interface{}, error)
  Del(key string) error
}
\end{lstlisting}
\end{minipage}
\caption{The Key-Value store interface.}
\end{example} \label{ex:KVStore}

Similar abstractions could (and will) be used to implement additional database types and functions. The key here is to create simple, higher-level interfaces on top of Textile Threads that simplifies dealing with events and data, but exposes the underlying event sourced data for customizing as needed. Developers should not have to learn a whole new set of terms and tools to take advantage of Textile's Threads capabilities. Other database abstractions include a no-sql style document store for storing and indexing arbitrary structs and/or JSON documents:

@@

Where \texttt{Indexable} can be satisfied by any structure with a \texttt{key} field and and \texttt{Query} is taken from the \texttt{go-datastore} interface library\footnote{\url{https://github.com/ipfs/go-datastore}} or similar.
Tables, feeds, counters, and other simple stores can also be built on Threads, and may require ORM-based solutions to support persistence to certain backing stores\footnote{In-memory stores are almost always implementable using compositions of basic data structures.}.
Each database style will be implemented as a standalone software library, allowing application developers to pick and choose the solution most useful to the application at hand.
Similarly, more advanced applications could be implemented using a combination of database types, or by examining the source code of these ``example'' libraries.

@@

\paragraph{Application State}

For application developers familiar with common state management issues, tools such as Redux provide an opinionated framework for making state mutations predictable by imposing certain restrictions on how and when updates can happen. In fact, Redux builds on concepts from CRQS and ES itself \cite{reduxMotivation}, and combines them with three core principals: 1) that the entire ``state' of an application should be stored in a single ``store'', 2) where the only way to change the state is to emit an ``action'' (or event), which 3) transforms the application state using a ``pure'' (i.e., zero side effects) reducer function (projection). In practice, Redux doesn't care how a developer persists the application state, which means a Thread's event store and read-only materialized views are excellent candidates. This focus on unidirectional data flow in perfectly in line with architecture proposed in this paper, and as such, a compatible Redux API will be built on top of Textile Threads.

>>>>

One of the most exciting aspects of user-siloed data is the the ability to build data-driven apps and services while users remain in control of their own data. Other projects have identified app silo-ed data as a problem [MISSING], and some have identified user-siloed data as a solution \cite{sambraSolidPlatformDecentralized2016}. However, none so far have provided a sufficiently interoperable protocol for scalable storage and transmission of user-siloed application data. This is the key to building great applications on user-siloed data.

@@

\begin{definition} (Command query responsibility segregation). Command query responsibility segregation or CQRS is a design pattern whereby reads and writes are separated into different models, using commands to write data, and queries to read data \cite{martinfowlerCQRS2011}. \end{definition}

@@

At a minimum, a node must link to its most immediate ancestor. However, links to older ancestors are often included as well to improve concurrency during traversal and verification \cite{meyerBamboo2019}.

@@

The use of well-established design patterns from the CQRS+ES (and related) literature means the Threads protocol can provide the foundation for many higher-level concepts and technologies. Like many CQRS-based systems, the ``public'' API for Threads revolves around the concept of Views, which are used to optimize queries on the local event store. A key feature of Views is they are completely disposable, because they can be entirely rebuilt from the source event store\footnote{Though see section \ref{sec:SnapshotsCompaction} for a discussion of Snapshots, which may be used to speed up Views.}. In Threads, Views are updated by subscribing to specific Event(s), and applying a Reducer function that is used to update the View state (which may be \emph{materialized} to local persistent storage). This simple subscription system provides a flexible framework for building complex View-driven application logic.

\subsubsection{Databse Views}\label{sec:DefaultViews}

The public API for a Thread already provides several features that you would expect when operating on a dataset or table within a database. Indeed, each Thread is defined by a unique ID (its Identity), and provides facilities for access control and permissions, networking, sharing, invites, and more. As such, it is not difficult to imagine a simplified API in which Threads are exposed via interfaces compatible with \emph{existing} datastores and/or communications systems.

The key here is to create simple, higher-level interfaces on top of Textile Threads that simplifies dealing with events and data, while still maintaining the power and flexibility of CQRS + ES. Developers should not have to learn a whole new set of terms and tools to take advantage of Textile's Threads capabilities. These simple, public-facing APIs will be comfortable to application developers looking to leverage highly-distributed systems that connect user-controlled data, with minimal configuration and maximum interoperability.

A database built on Threads would map database operations into Events. These Events would then mutate a View Model via its corresponding Reducer function, as per above. The actual View could be persisted or kept in-memory, and would be directly queried via a corresponding database-style APIs. Because Textile uses ES patterns, any delete-type operation would simply lead to an internal \emph{tombstone} Event, marking the database item for removal from the underlying materialized View.

\paragraph{Key/Value}

A key/value store built on Threads would ``map'' these database Events into key/value operations, such as \texttt{Put}, \texttt{Get}, and \texttt{Del}. These Events would then mutate a map-like View Model. To complete the database-like API, the entire API could be encapsulated in a database sturcture that satisfies a given interface (see for example, Figure \ref{ex:KVStore}).

\begin{example}
\begin{minipage}{.45\textwidth}
\begin{lstlisting}
type TextileKVStore interface {
  Put(key string, value Node) error
  Get(key string) (Node, error)
  Del(key string) error
}
\end{lstlisting}
\end{minipage}
\caption{The Key-Value store interface.}
\end{example} \label{ex:KVStore}

\paragraph{No-SQL}

Other database abstractions include a no-sql style document store for storing and indexing arbitrary structs and/or JSON documents. The interface for such as store, again built using a materialized View, might look like Figure \ref{ex:DocStore}, where \texttt{Indexable} could be satisfied by any structure with a \texttt{Key} field and \texttt{Query} might be taken from the \texttt{go-datastore} interface library\footnote{\url{https://github.com/ipfs/go-datastore}} or similar.

@@

\paragraph{Others}

Similar abstractions could (and will) be used to implement additional database types and functions. Tables, feeds, counters, and other simple stores can also be built on Threads, and may require ORM-based solutions to support persistence to certain backing stores\footnote{In-memory stores are almost always implementable using compositions of basic data structures.}. Each database style would be implemented as a standalone wrapper/software library, allowing application developers to pick and choose the solution most useful to the application at hand. Similarly, more advanced applications could be implemented using a combination of database types, or by examining the source code of these ``example'' libraries.

@@
